\MOC Server Pages:\\Web Publishing as Staged Computation

### 2007-04-16 paper/paper.tex
svn-r29: final? revision for SCP sent to Walid
darcs-hash:20070416223432-e5a07-285e748aea7af3d7a84fc751d85cfc848de4e182.gz

\section{Motivation}
\label{sec:intro}

Modern dynamic web sites support many features for user collaboration
and personalization.  To provide such services, web sites contain
custom computer programs, often written in one of a family of
programming languages that have grown up around (or been adapted for)
the web.

@@

This happens so often to sites featured on Slashdot---a
``news for nerds'' discussion site---that it has come to be
known as the \emph{Slashdot effect:} ``a site that might be
designed to handle a few hundred hits per day can suddenly
find itself handling that many a
second.''~\cite{turner03slashdot}~ Although some ISPs have
bandwidth limitations, Slashdot creator Rob Malda says most
sites that fail suffer from poor planning and architecture:
\begin{quote}
  The problem is, if you have any sort of complicated code
  on your page, that's what kills people. I generally think
  that sites are not dying because they fill their pipe, but
  because they have poorly written code.

  As a general rule, anybody who has a pretty good
  understanding of web design, they've done a good job of
  learning what information to cache. What information needs
  to be pregenerated. So when you're actually loading a
  page, even if it's a complicated page that looks dynamic
  and custom, on the back end of that, what they're really
  doing is putting together a bunch of puzzle pieces that
  have been pregenerated, and making the simplest, quickest
  decisions they possibly can.~\cite{turner03slashdot}
\end{quote}

Malda's observation points (informally) to the idea of \emph{staging}
the computation performed by a web service.  Indeed, web publishing is
an application area that is naturally staged:\footnote{Nevertheless,
  it seems the literature on meta-programming has not yet acknowledged
  web services~\cite{w3c-wsdl} as a potential application area,
  although \citet{sheard01accomp} does mention \emph{mobile code.}} %
\begin{enumerate}
\item Content (text, images, programs, etc.) created
  off-line is uploaded to the server---the \emph{publish}
  stage;

@@

\item The content is rendered within the user's
  browser---the \emph{display} stage.
\end{enumerate}
At \emph{each} stage there is an opportunity for computation
to take place.  Consider the example of a conference
calendar, such as the one illustrated in
figure~\ref{fig:confcal}.  After specifying your areas of
interest (perhaps using the ACM classifications), the server
delivers a table of matching conferences, with dates,
locations, deadlines, and links to conference web sites.
Events remain in the table until a few weeks after they
occur, but deadlines that have passed are marked in red.
You may click on any column header to change the sort order.
The next time you visit, the server remembers your
preferences.  Perhaps it even sends you email to remind you
of upcoming submission and registration deadlines.

@@

Now, how might this conference service be staged?  Can
anything be computed off-line (at the \emph{publish} stage)?
Yes!  Since this page is probably part of a much larger site
(whose structure does not change every day), the menus and
other navigation aids can be laid out in advance.  We will
not know which conferences to display until the user
presents some identification (in the form of a `cookie'),
but since the conference data change infrequently, it may
help to prepare the text of each row in advance.

@@

The \emph{serve} stage is well-served by the ``server page''
languages, including
JSP,\footnote{\url{http://java.sun.com/products/jsp/}}
ASP,\footnote{\url{http://www.asp.net/}} and
PHP.\footnote{\url{http://www.php.net/}} The Common Gateway Interface
(CGI)\footnote{\url{http://www.w3.org/CGI/}} addresses the
\emph{serve} stage, as do the embedded interpreters (such as
\texttt{mod\_perl} and \texttt{mod\_python} for Apache) that exist to
ameliorate some of the overhead of CGI.

There is, relatively, a paucity of languages that operate client-side
(\emph{display} stage), probably due to the difficulty of securing an
installed base of interpreters.  JavaScript and Java applets are
notable exceptions.

@@

Strictly speaking, these languages are not exclusively confined to the
stages that I have indicated: Javascript can be run client-side, PHP
can be run off-line, and so on.  Nevertheless, migrating code between
stages is hard, and the need for quoting and persistence are
practically show-stoppers.  For comparison purposes (further described
in section~\ref{sec:perf}), we staged some code using PHP.  To achieve
persistence of composite variables from one stage to the next, it
contains gems like this:\label{php-gem}
\begin{verbatim}
    <?= "<? \$list = unserialize(\"".
        addcslashes(serialize($list),'"').
        "\"); ?>\n" ?>
\end{verbatim}
where the \texttt{serialize} library function occurs in stage one and
the \texttt{unserialize} in stage two.  Notice that the \verb+$+
preceding the first occurrence of \verb+list+ is quoted, but the
second occurrence is not.  The \texttt{addcslashes} function is needed
in case the serialized representation contains special characters
(such as double quotes or newlines) that would be misinterpreted by
the PHP parser in the next stage.  Interpreting the stage-one program
guarantees nothing about the well-formedness of the stage-two program
(generated as a string).

@@

A very common use of code blocks is to print out (i.e., send to the
browser) the result of evaluating some expression.  The syntax
\cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must have
type \cd{string}.  Alternatively, messages may be formatted with
\cd{printf} by placing the format string immediately after the code
delimiter.\footnote{In a departure from the standard OCaml syntax for
  \cd{printf}, the arguments are comma-delimited.  This way, fewer
  parentheses are required when using the escape or lift operators on
  the arguments.}
%%
\lgrindfile{pi.meta}
The output is:
\begin{quote}
  ........ $\pi$ $\div$ 008 is 0.3927
\end{quote}

@@

\MOC augments OCaml with just three annotations, to indicate how
programs are to be staged.  Brackets \cd{\Mbegincode like
  these\Mendcode} construct future-stage computation.  The code within
is not executed \emph{now,} but just returned, as a code value that can be
run later.

Within brackets, the splice or escape operator \cd{\Msplice e} may
appear.  It interrupts the code construction to execute the expression
\cd{e} (in the current stage) and splice its result into the
future-stage computation.  Thus, \cd{e} is required to return code of
the proper type.

@@

The \cd{a b c} represent publish-stage arguments (the names are
specified in the \cd{pragma}), whereas \cd{req} and \cd{puts} are
(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
details, including the headers and query arguments.  \cd{puts} is a
function, provided by the server, to transmit text across the network
to the user's browser.  
%%
The library function \cd{Request.arg} of the type \cd{request\to
  string\to string option} looks up the string value of the request
parameter with a given name.  Since \cd{req} and \cd{puts} are
serve-stage arguments, it is incorrect to use them in the
\emph{publish} (first) stage, and indeed the \MOC type system prevents
this.

A few finer points to note: the translator discards newlines that
immediately follow code blocks (otherwise,
figure~\ref{fig:trans.ml} would be dotted with %
\cd{puts}~\verb+"\n"+ statements).  It automatically appends a
semi-colon or the {\KWfont in} keyword to code blocks, as required
(see, for example, lines 7, 11, and 12 in figure~\ref{fig:trans.ml}.)
%%
Finally, because the translator partially parses the OCaml code
blocks, it is not confused by code delimiters within strings and
comments, nor by other uses of \la{} and \ra{} as operators.

\subsection{Staged code blocks}
Now that we understand how the server page is assembled into a staged
program, the effects of adding \MOC staging operators to our pages
should be predictable.  Below is another example using \cd{stat}, this
time to display the size of the server log file.
With the serve-stage argument \texttt{unit}, we can specify whether
the size should be expressed in bytes (the default), kilobytes, etc.
\lgrindfile{size.meta}%

To stage this, the \cd{stat} and \cd{float\_of\_int} calls could be
lifted into the declaration block, as with the permissions example.
Furthermore, we may also use \cd{lift} and the splice operator to
perform the divisions in advance, even though they are underneath the
{\KWfont match} (which cannot happen until the serve stage):
\lgrindfile{size2.meta}%

@@

We will see examples of most of these blocks in section~\ref{sec:eg}
and appendix~\ref{sec:dir.meta}.

@@

\MOC server pages support publish-stage arguments, as demonstrated by
the identifiers \cd{a b c} in figures~\ref{fig:trans.meta}
and~\ref{fig:trans.ml}.  The programmer specifies the pattern to be
used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
may appear anywhere in the code.

With publish-stage arguments, the page may be instantiated in
countless ways.  Continuing the count-down example, we could map the
URI \id{/longcount} to a code block generated with an actual argument
of 99, while \id{/shortcount} refers to code with the actual
argument 9.  Both are generated from the same server page.

@@

To explore the expressiveness of this design, we now look at a series
of web services, organized to comprise a small web site.\footnote{A
  demo is available at \url{http://comsci.liu.edu/metaserv/}} To give
the services a similar look, we developed a site-wide style sheet, and
an OCaml function to generate a navigation bar from a hierarchical
list of page titles and links.  
%%
\mylst{navspec-sig.ml}{These functions help define the standard site
  layout.}
%%
The page layout functions in figure~\ref{fig:navspec-sig.ml} may be
invoked at any stage.  Naturally, if the page title includes a dynamic
argument, then \cd{preamble} will have to be delayed until the serve
stage.

@@

In section~\ref{sec:perf}, we will measure the impact of staging on
the scalability of this application.  To derive the un-staged version
for comparison, we simply convert \cd{\Mquote{\Mtilde}} and
\cd{\Mquote{!}} blocks into plain \cd{\Mquote{}} and remove all other
brackets and escapes from the code in figure~\ref{fig:power.meta}.

@@

Now we consider a more substantial example.
Many web servers can be configured to permit clients to browse
directories.  The web server generates, on the fly, an HTML page
containing the names and attributes of (and hyperlinks to) all
the files in the directory.  In Apache, the
\texttt{mod\_autoindex} module provides this feature.
%%
ViewCVS\footnote{\url{http://viewcvs.sourceforge.net/}} is a more
complex example of the same idea; it allows remote users to browse a
CVS repository with a standard web browser.

@@

The \cd{FileHandler} takes a file system path as a parameter, and then
interprets URIs as file names relative to that path.  If such a file
exists, it sends it verbatim to the client.  If not, it passes
the responsibility on to the next handler in the chain.

The \cd{CodeHandler} is provided with a dictionary to map URIs to code
values.  It looks up the URI in the dictionary, and if a match is
found it invokes that code.  For now, the map is hard-coded at build
time and the code values are already in memory when the server starts
accepting requests.  This is because \MOC is currently not capable of
reading and writing code values to a file.

Here are the types of the page code and the map, along with signatures
for the \cd{CodeHandler} (and two variants):

\lgrindfile{codeHndl.mli}
%%
\cd{page} is the type of the code that is constructed by each server
page; recall that the page takes two serve-time arguments: one
encapsulating the HTTP request, the other is the \cd{puts} function
for sending text back to the client.

Values of type \cd{map} tell the server which pages are mapped to
which URIs.  Ideally, it would be defined as \cd{page code
  StringMap.t}.  This is not currently possible: the \cd{page code}
needs to be run, but in the current implementation of \MOC, the
\Mrun{} operator cannot be used from within functions.  Passing around
code values themselves has limited utility because they must remain
polymorphic to run them, and OCaml does not yet support rank-2
polymorphism.  (Although recent work by \citet{garrigue99semi} leads
in that direction.)  Our work-around involves mapping the URI instead
to a pair containing a function (defined at the top level) to generate
the page (by running the first stage and then using the \Mrun{}
operator on the generated code) and a ref cell where the resulting
page is cached.

@@

For the ``OCaml CodeHandler'' series, the same set of files were
treated as \MOC Server Pages, and thus translated (in advance) into
one big \texttt{puts} statement, to be executed (as byte code) by the
\texttt{CodeHandler} module.  The only reason the code beats the (byte
code) FileHandler in the beginning is that all code pages are already
loaded into the server's memory on startup (to work around a
limitation of the prototype---see section~\ref{sec:impl}), but the
FileHandler must read files from disk each time.

@@

At the suggestion of an anonymous reviewer, we also tried staging
using higher-order functions in OCaml instead of the code splicing
features of \MOC.  The relavant fragment of code is shown in
figure~\ref{fig:powerhof.ml}.  Using continuation-passing style, it
recursively builds up a closure that represents the computation
required to compute any number to the power \cd n.

@@

Besides the staging factor, two implementation languages are compared:
we implemented the same functionality in PHP.  The `staged PHP' version must
be run first from the command line; this outputs a PHP script
which is then run by the server.  With PHP we must, as noted in
section~\ref{sec:intro}, manage the quoting and persistence 
by hand.  (This program is the source of the horrible
\texttt{serialize}/\texttt{unserialize} code shown on
page~\pageref{php-gem}.

Directory browsing was the most realistic of the examples, and the
benefit of staging is crystal clear.  In browsing a directory of 64
files, the unstaged programs barely answered 18 requests per second.
They would certainly succumb to the Slashdot effect!  (And compilation
to native code made essentially no difference.)

This is exactly the kind of page where the real work needs to be done
in advance.  But that doesn't mean it needs to be a completely static
page, either.  By carefully staging the computation, we gathered the
file information in advance, yet still filtered and sorted the results
on demand.  The staged \MOC directory browser answered more than 550
requests (for the same 64-file directory) per second.

@@

@@

One of the major shortcomings of our implementation is that all the
pages must be loaded into memory when the server starts.  Ideally, a
separate tool would translate the server page source directly to
byte-code, stored in an image on disk.  Then the server would map URIs
to these byte-code files, loading and executing them on demand.
Whenever we want to re-run the publish stage, we could do so
independently of the server.  Two limitations of the \MOC
implementation itself appear to stand in the way of this ideal.
First, code values cannot at present be written to or read from files.
Second, code values remain polymorphic until they are \emph{run} with
the \Mrun{} operator.  Although they may be passed to functions and
stored in data structures, code values \emph{cannot be run} from
within functions.  That is, the \Mrun{} operator can only be used at
the top level (until OCaml supports rank-2 polymorphism), which
severely restricts the kinds of multi-stage programs one can write.

Another interesting avenue is to incorporate work on
\emph{offshoring}~\cite{eckhardt04offshoring}.  An alternative
\emph{run} construct translates generated code blocks to lower-level
languages for improved performance.  In the domain of web services, it
is conceivable that a \MOC server page could then generate (via
offshoring) a specialized server page in a more mainstream language,
such as PHP or JSP.

@@

Many thanks to the anonymous reviewers for the \MOC workshop, and for
this article, for many helpful and enlightening recommendations on
both the work and the presentation.  Thanks to Walid Taha for
organizing the workshop, and to the rest of the \MOC team for their
development efforts.

\bibliographystyle{abbrvnat}
\bibliography{refs,../../cal}
\appendix
%\twocolumn
\section{The directory browser}
\label{sec:dir.meta}
%\raggedright
\setcodefonts{11pt}{10pt}
\LGnuminterval=5000
\LGindent=8pt
\gdef\lgcommentstart{\end{lgrind}\vskip-3pt}
\def\lgcommentend{\vskip.5\baselineskip\begin{lgrind}}

>>>>

\section{Motivation}
\label{sec:intro}
Modern dynamic web sites support many features for user collaboration
and personalization.  To provide such services, web sites contain
custom computer programs, often written in one of a family of
programming languages that have grown up around (or been adapted for)
the web.

@@

This happens so often to sites featured on Slashdot---a
``news for nerds'' discussion site---that it has come to be
known as the \emph{Slashdot effect:} ``a site that might be
designed to handle a few hundred hits per day can suddenly
find itself handling that many a
second.''~\cite{turner03slashdot}~ Although some ISPs have
bandwidth limitations, Slashdot creator Rob Malda says most
sites that fail suffer from poor planning and architecture:
\begin{quote}
  Anybody who has a pretty good
  understanding of web design [has] done a good job of
  learning what information to cache [and] what needs
  to be pre-generated.  So when you're actually loading a
  page, even if it's a complicated page that looks dynamic
  and custom, on the back end of that, what they're really
  doing is putting together a bunch of puzzle pieces that
  have been pre-generated, and making the simplest, quickest
  decisions they possibly can.~\cite{turner03slashdot}
\end{quote}

Malda's observation points (informally) to the idea of \emph{staging}
the computation performed by a web service.  Indeed, web publishing is
an application area that is naturally staged:\footnote{The literature
on meta-programming has yet to acknowledge
  web services~\cite{w3c-wsdl} as a potential application area,
  although \citet{sheard01accomp} mentions \emph{mobile code.}
  Analysis of other related work is in section~\ref{sec:related}.} %
\begin{enumerate}
\item Content (text, images, programs, etc.) created
  off-line is uploaded to the server---the \emph{publish}
  stage;

@@

\item The content is rendered within the user's browser---the
  \emph{display} stage.\footnote{Increasingly in modern web
    applications, the rendered content is
    \emph{interactive}---responsive to user input without a round-trip
    to the server.}
\end{enumerate}
At \emph{each} stage there is an opportunity for computation to take
place.\footnote{\citet{normark02programmatic} recognized these three
stages as binding times, calling them 
\emph{generated, calculated,} and \emph{dynamic} documents,
respectively.}
Consider the example of a conference calendar, such as the one
illustrated in figure~\ref{fig:confcal}.  After specifying your areas
of interest (perhaps using the ACM classifications), the server
delivers a table of matching conferences, with dates, locations,
deadlines, and links to conference web sites.  Events remain in the
table until a few weeks after they occur, but deadlines that have
passed are marked in red.  You may click on any column header to
change the sort order.  The next time you visit, the server remembers
your preferences.  Perhaps it even sends you email to remind you of
upcoming submission and registration deadlines.

@@

Now, how might this conference service be staged?  Can
anything be computed off-line (at the \emph{publish} stage)?
Yes: since this page is probably part of a much larger site
(whose structure does not change every day), the menus and
other navigation aids can be laid out in advance.  We will
not know which conferences to display until the user
presents some identification (in the form of a `cookie'),
but since the conference data change infrequently, it may
help to prepare the text of each row in advance.

@@

The \emph{serve} stage is well-served by the ``server page''
languages, including
JSP,
ASP, and
PHP.  The Common Gateway Interface
(CGI)\footnote{\url{http://www.w3.org/CGI/}} addresses the
\emph{serve} stage, as do the embedded interpreters (such as
\texttt{mod\_perl} and \texttt{mod\_python} for Apache) that exist to
ameliorate some of the overhead of CGI.

There is, relatively, a paucity of languages that operate client-side
(\emph{display} stage), probably due to the difficulty of securing an
installed base of interpreters.  JavaScript, Java, and Flash applets are
notable exceptions.

@@

Strictly speaking, these languages are not exclusively confined to the
stages that I have indicated: Javascript can be run server-side, PHP
can be run off-line, and so on.  Nevertheless, migrating code between
stages is hard, and the need for quoting and persistence are
practically show-stoppers.  For comparison purposes (further described
in section~\ref{sec:perf}), we staged some code using PHP.  To achieve
persistence of composite variables from one stage to the next, it
contains gems like this:\label{php-gem}
\begin{verbatim}
    <?= "<? \$list = unserialize(\"".
        addcslashes(serialize($list),'"').
        "\"); ?>\n" ?>
\end{verbatim}
where the \texttt{serialize} library function occurs in stage one and
the \texttt{unserialize} in stage two.  Notice that the \verb+$+
preceding the first occurrence of \verb+list+ is quoted, but the
second occurrence is not.  The \texttt{addcslashes} function is needed
in case the serialized representation contains special characters
(such as double quotes or newlines) that would be misinterpreted by
the PHP parser in the next stage.  Interpreting the stage-one program
guarantees nothing about the well-formedness of the stage-two program
(generated as a string).

@@

In this example and throughout this paper, a {\BGfont sans-serif}
font is used for embedded \MOC code, with {\KWfont bold sans} reserved
for keywords and code delimiters.  A \texttt{typewriter} font is used
for \MOC character strings.  The regular serif font is used for plain
text and HTML within the server page, and for comments within the \MOC
code blocks.

A very common use of code blocks is to print out (i.e., send to the
browser) the result of evaluating some expression.  The syntax
\cd{\Mquote= e\Munquote} is designated for this task; \cd{e} must have
type \cd{string}.  Alternatively, messages may be formatted with
\cd{sprintf} by placing the format string immediately after the code
delimiter.\footnote{In a departure from the standard OCaml syntax for
  \cd{sprintf}, the arguments are comma-delimited.  This way, fewer
  parentheses are required when using the escape or lift operators on
  the arguments.}
%%
\lgrindfile{pi.meta}
The output is:
\begin{quote}
  ........ $\pi$ $\div$ 008 is 0.3927
\end{quote}

@@

\MOC augments OCaml with just three annotations, to indicate how
programs are to be staged.  Brackets \cd{\Mbegincode e\Mendcode}
construct future-stage computation.  The code within is not executed
in the current stage of computation, but just returned as a code value
that can be run later.

Within brackets, the splice or escape operator \cd{\Msplice e} may
appear.  It interrupts the code construction to evaluate the expression
\cd{e} (in the current stage) and splice its result into the
future-stage computation.  Thus, \cd{e} is required to evaluate to code of
the proper type.

@@

The \cd{a b c} represent publish-stage arguments (the names are
specified with \cd{pragma args}), whereas \cd{req} and \cd{puts} are
(fixed) serve-stage arguments.  \cd{req} encapsulates the HTTP request
details, including the headers and query arguments.  \cd{puts} is a
function, provided by the server, to transmit text across the network
to the user's browser.  
%%
The library function \cd{Request.arg} of the type \cd{request\to
  string\to string option} looks up the string value of the request
parameter with a given name.  The \cd{option} type constructor permits
the function to return \cd{None} if there is no matching parameter in
the HTTP request.  On line 6, \cd{arg} is defined as a short-cut for
retrieving a parameter by name.  Since \cd{req} and \cd{puts} are
serve-stage arguments, it is incorrect to use them in the
\emph{publish} (first) stage, and indeed the \MOC type system prevents
this.  These, along with the \cd{lift} function defined near the top,
are essentially primitives from the point of view of the server page
code.%%
%%
\footnote{Here are a few finer points about the translator: it discards
newlines that immediately follow code blocks (otherwise,
figure~\ref{fig:trans.ml} would be dotted with
\cd{puts}~\texttt{"\char"5C{}n"} 
statements).  It automatically appends a semi-colon or the {\KWfont
in} keyword to code blocks, as required (see, for example, lines 7,
11, and 12 in figure~\ref{fig:trans.ml}.)  Finally, because the
translator partially parses the OCaml code blocks, it is not confused
by code delimiters within strings and comments, nor by other uses of
\la{} and \ra{} as operators.}

\subsection{Staged code blocks}
Now that we understand how the server page is assembled into a staged
program, the effects of adding \MOC staging operators to our pages
should be predictable.  Below is another example using \cd{stat}, this
time to display the size of some text file on the server.
With the serve-stage argument \texttt{unit}, we can specify whether
the size should be expressed in bytes (the default), kilobytes, etc.
\lgrindfile{size.meta}%

If this text file does not change frequently (and reporting 
outdated information is no problem), the \cd{stat} and
\cd{float\_of\_int} calls could be lifted into the declaration block,
as with the permissions example.  Furthermore, we may also use
\cd{lift} and the splice operator to perform the divisions in advance,
even though they are underneath the {\KWfont match} (which cannot
happen until the serve stage): \lgrindfile{size2.meta}%

@@

In these cases, the expressions should not have code types: \cd{a}
simply has type \cd{unit}, \cd{b} has type string, etc.  This is the
only difference between the \cd{\Mtilde} and \cd{!} variants; both are
evaluated in the first stage.  We will see examples of most of these
blocks in section~\ref{sec:eg} and appendix~\ref{sec:dir.meta}.

@@

\MOC server pages support publish-stage arguments, as demonstrated by
the identifiers \cd{a b c} in figures~\ref{fig:trans.meta}
and~\ref{fig:trans.ml}.  The programmer specifies the pattern to be
used with a \cd{\Mquote{}pragma args ...\Munquote} declaration, which
may appear anywhere in the code.  Any identifiers following \cd{args}
in the \cd{pragma} declaration will become publish-stage parameters.

With publish-stage parameters, the page may be instantiated in
countless ways.  Continuing the count-down example, we could map the
URI \id{/longcount} to a code block generated with an argument of 99,
while \id{/shortcount} refers to code with the argument 9.  Both are
generated from the same server page.

@@

To explore the expressiveness of this design, we now look at a series
of web services, organized to comprise a small web site.  To give
the services a similar look, we developed a site-wide style sheet, and
an OCaml function to generate a navigation bar from a hierarchical
list of page titles and links.  
%%
\mylst{navspec-sig.ml}{These functions help define the standard site
  layout.}
%%
The page layout functions in figure~\ref{fig:navspec-sig.ml} may be
invoked at any stage.  Naturally, if the page title includes a dynamic
argument, then \cd{preamble} will have to be delayed until the serve
stage.

@@

In section~\ref{sec:perf}, we will measure the impact of staging on
the scalability of this application.  To derive the un-staged version
for comparison, we simply replace \cd{\Mquote{\Mtilde}} and
\cd{\Mquote{!}} blocks with plain \cd{\Mquote{}} and remove all other
brackets and escapes from the code in figure~\ref{fig:power.meta}.

@@

Now we consider a more substantial example.
Many web servers can be configured to permit clients to browse
directories.  The web server generates, on the fly, an HTML page
containing the names and attributes of (and hyperlinks to) all
the files in whatever directory is specified by the URL.  In Apache, the
\texttt{mod\_autoindex} module provides this feature.
%%
ViewCVS is a more
complex example of the same idea; it allows remote users to browse a
CVS repository with a standard web browser.

@@

The \cd{FileHandler} takes a file system path as a parameter, and then
interprets each URI as a file name relative to that path.  If such a
file exists, it sends it verbatim to the client.  If not, it passes
the responsibility on to the next handler in the chain.

The \cd{CodeHandler} is provided with a dictionary to map URIs to code
values.  It looks up the URI in the dictionary, and if a match is
found it invokes that code.  For now, the map is hard-coded at build
time and the code values are already in memory when the server starts
accepting requests.  This is because \MOC does not currently permit
programs to read and write code values to a file.

Here are the types of the page code and the map, along with the
signature for the \cd{CodeHandler}:

\lgrindfile{codeHndl.mli}
%%
\cd{page} is the type of the code that is constructed by each server
page; recall that the page takes two serve-time parameters: one
encapsulating the HTTP request, the other is the \cd{puts} function
for sending text back to the client.

Data structures of type \cd{map} tell the server which pages are
mapped to which URIs.  The range of the map is a pair: the first
component (of type \cd{unit\to page}) is a function that will re-run
stage one computation (using the \Mrun{} operator internally); the
second component (of type \cd{page ref}) is a cell where the latest
page is cached.

The \cd{unit\to page} formulation is a work-around for what might
otherwise be expressed as \cd{page code}, and run from within the
\cd{CodeHandler}.  Unfortunately, code values in MetaOCaml must remain
polymorphic to run them; the type is actually \cd{('a, page) code},
for all \cd{'a}.  This is traditionally problematic in ML: we can
define a polymorphic data structure, but not a data structure
containing polymorphic values.  The work-around we used captures the
polymorphic code value in a closure, which is then added to the map.
Another possibility is to use the limited form of explicit
polymorphism supported in OCaml, with the syntax \cd{\{f : 'a. ('a,
page) code\}}.\footnote{Thanks to an anonymous reviewer for pointing
this out.}  To use code values more directly, we would need rank-2
polymorphism; work by \citet{garrigue99semi} is headed in that
direction.

@@

For the ``\MOC CodeHandler'' series, the same set of files were
treated as \MOC Server Pages, and thus translated (in advance) into
one big \texttt{puts} statement, to be executed (as byte code) by the
\texttt{CodeHandler} module.  The only reason the code beats the (byte
code) FileHandler in the beginning is that all code pages are already
loaded into the server's memory on startup (to work around a
limitation of the prototype---see section~\ref{sec:impl}), but the
FileHandler must read files from disk each time.

@@

At the suggestion of an anonymous reviewer, we also tried staging
using higher-order functions in OCaml instead of the code splicing
features of \MOC.  The relevant fragment of code is shown in
figure~\ref{fig:powerhof.ml}.  It
recursively builds up a closure that represents the computation
required to compute any number to the power \cd n.

@@

Besides the staging factor, two implementation languages are compared:
we implemented the same functionality in PHP.  The `staged PHP' version must
be run first from the command line; this outputs a PHP script
which is then run by the server.  With PHP we must, as noted in
section~\ref{sec:intro}, manage the quoting and persistence 
by hand.  (This program is the source of the horrible
\texttt{serialize}/\texttt{unserialize} code shown on
page~\pageref{php-gem}.)

Directory browsing was the most realistic of the examples, and the
benefit of staging is crystal clear.  In browsing a directory of 64
files, the unstaged programs barely answered 18 requests per second.
They would certainly succumb to the Slashdot effect---and compilation
to native code made essentially no difference.

This is exactly the kind of page where the real work needs to be done
in advance.  But that does not mean it needs to be a completely static
page, either.  By carefully staging the computation, we gathered the
file information in advance, yet still filtered and sorted the results
on demand.  The staged \MOC directory browser answered more than 550
requests (for the same 64-file directory) per second.

@@

There is much related work on using various features of modern
programming languages to implement sophisticated web services.
\citet{normark02programmatic} proposed writing web pages in Scheme
using the Lisp Abstracted Markup Language (LAML), which essentially
represents HTML documents as S-expressions.  He distinguished three
different binding times, when the Scheme program could be evaluated:
off-line, page access time, or browse time (client-side).  These
correspond precisely to the three stages we identified in section 1,
but his programs did not \emph{transcend} different stages.  It is
possible to implement staged programming in Scheme using
\cd{quasiquote}, \cd{unquote}, and \cd{eval}~\cite{davies01modal};
this would lead to a similar capability, save for the difference
between static and dynamic checking of the generated code.

\citet{queinnec00influence} and \citet{hughes00generalising} observed
that multi-page web services could be implemented more naturally using
continuations and \cd{call/cc} (essentially, by treating the server
and user as coroutines).  This way, we can treat the entire service as
one program---that suspends itself while waiting for user
input---instead of developing each page as a separate program.  This
is a valuable technique, and appears to be orthogonal to staging;
although it would be worth implementing both together to determine if
there are unforeseen interactions.

\citet{graunke01programming,graunke03modeling} took up this idea and
mixed it with other language features: first-class modules, preemptive
threads, and \emph{custodians,} for managing resource consumption.
The result is a server that achieves lower overhead for dynamic
services compared to Apache CGI (although the overhead of CGI is
avoided by most modern dynamic services, by embedding interpreters for
JSP, PHP, or similar in the server).  \citet{matthews04automatically}
compile such direct-style interactive programs into CGI-style scripts.

@@

One of the major shortcomings of our implementation is that all the
pages must be loaded into memory when the server starts.  Ideally, a
separate tool would translate the server page source directly to
byte-code, stored in an image on disk.  Then the server would map URIs
to these byte-code files, loading and executing them on demand.
Whenever we want to re-run the publish stage, we could do so
independently of the server.

Another interesting avenue is to incorporate work on
\emph{offshoring}~\cite{eckhardt04offshoring}.  An alternative
\emph{run} construct translates generated code blocks to lower-level
languages for improved performance.  In the domain of web services, it
is conceivable that a \MOC server page could then generate (via
offshoring) a specialized server page in a more mainstream language,
such as PHP, JSP, or---for client-side computation---JavaScript.

@@

Sincere thanks to the anonymous reviewers, for many helpful and
enlightening recommendations on both the work and the presentation.
Thanks to Walid Taha for organizing the \MOC workshop, and to the rest
of the \MOC team for their development efforts.

\bibliographystyle{abbrvnat}
\bibliography{refs,../../bib/library}
\appendix
%\twocolumn
\section{The directory browser}
\label{sec:dir.meta}
%\raggedright
\setcodefonts{11pt}{10pt}
\LGnuminterval=5000
\LGindent=8pt
\gdef\lgcommentstart{\end{lgrind}\vskip-3pt}
\def\lgcommentend{\vskip.5\baselineskip\begin{lgrind}}
